The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2015-01-14 articles/2014/MWA-Radio-Environment/response-to-referee.txt

Dear referee and editor,

We would like to thank the referee for his comments on our article. We agree with most suggestions by the referee. For a few comments by the referee, we think no changes were required or we have made a change slightly different as suggested. These cases are detailed below and we think are minor differences. Other than those, all suggestions were implemented, and we think the paper is ready for publication. All changes have been marked in the article with a red/bold font.

Page 7
"Because the AOFLAGGER determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig. 1), it is insensitive to RFI covering such a broad spectrum."

Referee: "Because of the increase in spread-spectrum transmissions, this is a quite serious weakness in AOFLAGGER. The process described in the next paragraph might be included into the pipeline in a generic way to detect any spread-spectrum RFI."

and from the decision letter:
"AOFLAGGER fails to detect what is becoming an increasingly troublesome source of RFI: persistent spread-spectrum signals. A methodology for supplementing AOFLAGGER using COTTER statistics is suggested, but some discussion of how AOFLAGGER may be modified to handle this class of RFI would be useful. "

Answer: The extra flagging step runs fast compared to AOFlagger (which is described in the paper). It is therefore not necessary to include this step inside Cotter or AOFlagger -- a reason for inclusion in the initial pre-processing pipeline is when it needs the data at high resolution, but that's not the case. While this RFI is serious, it's not a serious weakness in the flagger that such RFI is not immediately flagged; the initial flagger is aimed at flagging out 'high-resolution' RFI that occupies a few channels or timesteps, and a second flagging step can recognize broadband RFI at a lower data resolution. We think this is described well enough in the paper and no further changes are required.

Page 8
Sentence: "[..] and the exponent alpha is typically -1.5 to -1.6"

@@

Referee: "Presumably these nominal values of alpha are determined empirically?  This should be stated explicitly."

Answer: In Offringa et al (2013b), this is both empirically found as well as mathematically derived to be true for uniform distributions of sources with a given average transmitter height. We think explaining this completely is not necessary, so instead we now refer to the paper: "the exponent alpha is found by Offringa et al (2013b) to be typically -1.5 to -1.6."
and explain the uniformness property: "Using LOFAR, Offringa et al. (2013b) show that when a uniform spatial distribution with sufficient interfering transmitters is observed, the distribution of the visibility brightness will have a power-law tail [..]"

Page 9
"[..] a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power on all baselines is added together. We currently do not know what their origin is"

@@

Referee: "Such seemingly random impulsive RFI is ubiquitous - probably caused by ESD "sparking" and relay contacts."

Answer: We agree with the referee, but rather don't speculate to what is the cause.

Page 10, Figure 6.
Referee: "Are the centre channels of the PFB bands also flagged?  Otherwise, what are narrow horizontal lines?"

Answer: This is explained in the caption below the image (which is placed at a somewhat incorrect position by the PASA-generated PDF, so that might have been a bit confusing) : "[..] The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively."

Page 12
"In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia A and Cygnus A are only visible at low elevations."

Referee: "The southern hemisphere has its own strong discrete sources - are these not evident?"

Answer: At these frequencies, Cyg A and Cas A are almost an order of magnitude larger (~5-10 KJy) than the strongest Southern sources such as M1 (~1 KJy).

### 2015-01-14 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Referee report implemented

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose \editmark{unsurmountable} issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} conclude that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth \editmark{into 48 equal sub-bands and dividing each sub-band} by its Winsorized standard deviation \editmark{(see \citet{variance-estimates} for an explanation of Winsorized statistics}). This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only used during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but it was necessary to apply this correction for the data used here.

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. The MWA has a 30.72~MHz instantaneous bandwidth, and \editmark{observing at two different central frequencies is therefore required to cover the desired EoR range. Observations are centred on either 154.2 or 182.4~MHz, with an overlap of 2.6~MHz. The bands covered by these central frequencies} will be referred to as the EoR low and high bands, respectively. For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary \editmark{tied-array} beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz. The GLEAM observations are made in drift-scan mode, with the 5 bands rotated through in sequence on a 2-minute cadence.

@@

Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section. \editmark{RFI occupancy is calculated as the percentage of discrete visibilities within the frequency range that are detected as RFI by the flagger.} The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

@@

In the ideal case, the residual visibility distribution would show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). The residual curves in Fig.~\ref{fig:lognlogs} \editmark{do not fall off with RFI power as quickly} as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes. This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam. The EoR-high distribution has some extra features in its tails that are not smooth. Further analyses showed that these are also caused by the Galaxy, causing a few short baselines to observe samples with high amplitudes. Visual inspection of snapshots with a non-smooth residual distribution tail did not show residual RFI in such sets.

@@

Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz). This is the worst \editmark{example of} contamination in our data sets by this transmitter, and there are snapshots in which the transmitter is not visible at all. This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions. While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot. In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is shown but is zoomed in on a briefly-observed RFI event at 150.17 MHz. This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.

@@

\section*{Acknowledgements}
This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the U.S. National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.

>>>>

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose \editmark{unsurmountable} issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} \editmark{conclude} that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth \editmark{into 48 equal sub-bands and dividing each sub-band} by its Winsorized standard deviation \editmark{(see \citet{variance-estimates} for an explanation of Winsorized statistics}). This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only \editmark{used} during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but it was necessary to apply this correction for the data used here.

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. The MWA has a 30.72-MHz instantaneous bandwidth, \editmark{which makes it necessary to observe at two different frequencies to cover the desired EoR bandwidth. An observing night is centred on either 154.2 or 182.4~MHz, which results in an overlap of 2.6~MHz. The bands covered by these central frequencies} will be referred to as the EoR low and high bands, respectively. For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary \editmark{tied-array} beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz. The GLEAM observations are made in drift-scan mode, with the 5 bands rotated through in sequence on a 2-minute cadence.

@@

Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section. \editmark{RFI occupancy is calculated as the percentage of discrete visibilities that are detected as RFI by the flagger at the resolution of the correlator output.} The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

@@

In the ideal case, the residual visibility distribution would show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). The residual curves in Fig.~\ref{fig:lognlogs} \editmark{do not fall off with RFI power as quickly} as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes. This excess is caused by the \editmark{variable} nature of the data, for example because the noise level increases when the Galaxy goes through the beam. The EoR-high distribution has some extra features in its tails that are not smooth. Further analyses showed that these are also caused by the Galaxy, causing a few short baselines to observe samples with high amplitudes. Visual inspection of snapshots with a non-smooth residual distribution tail did not show residual RFI in such sets.

@@

Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz). This is the worst \editmark{example of} contamination in our data sets by this transmitter, and there are \editmark{snapshots} in which the transmitter is not visible at all. This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions. While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot. In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is shown but is zoomed in on a briefly-observed RFI event at 150.17 MHz. This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.

@@

\section*{Acknowledgements}
\editmark{This scientific work makes use of the NCI National Facility in Canberra, Australia, which is supported by the Australian Commonwealth Government. }

This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the U.S. National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.
